The SNIK Ontology of Hospital Information Management

### 2019-03-01 paper.tex
Clean main template and correct todonotes display.

% please run ``latex paper'' and then ``makeglossaries paper'' from the command line to create the glossaries files
% add. options: [seceqn,secthm,crcready,onecolumn]
\documentclass[sw]{iosart2x}

%\usepackage{dcolumn}
%\usepackage{endnotes}

\usepackage{booktabs}
\usepackage{csquotes}
\usepackage{graphicx}
\usepackage{listings}
\lstset{language=SPARQL,breaklines=true}

\usepackage{cleveref}

@@

% Two or more authors:
\author[A]{\inits {K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@imise.uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}},
\author[A]{\fnms{Franziska} \snm{Jahn}\ead[label=e2]{konrad.hoeffner@imise.uni-leipzig.de}}
\author[A]{\fnms{Birgit} \snm{Schneider}\ead[label=e3]{birgit.schneider@imise.uni-leipzig.de}}
\author[A]{\fnms{Anna} \snm{Lörke}\ead[label=e4]{anna.loerke@imise.uni-leipzig.de}}
\author[A]{\fnms{Thomas} \snm{Pause}\ead[label=e5]{thomas.pause@imise.uni-leipzig.de}}
\author[A]{\fnms{Alfred} \snm{Winter}\ead[label=e6]{alfred.winter@imise.uni-leipzig.de}}
\runningauthor{}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE),
\orgname{University of Leipzig}, \cny{Germany}\printead[presep={}}}]{e1,e2,e3,e4,e5,e6}}
%Medical Informatics, Management of Health Information Systemsi
%Härtelstraße 16--18, D-04107 Leipzig
E-mail: \{khoeffner,fjahn,alfred\}@imise.informatik.uni-leipzig.de}
\begin{abstract}
Abstract text.
\end{abstract}
\begin{keyword}
\kwd{Hospital Information Management}
\end{keyword}
\end{frontmatter}
%%%%%%%%%%% The article body starts:
\section{Introduction}\label{sec:introduction}
A W3C design issue~\cite{puttinggovernmentdataonline} motivates making government data available online as Linked Data for three reasons:
\enquote{1) Increasing citizen awareness of government functions to enable greater accountability; 2) Contributing valuable information about the world; and 3) Enabling the government, the country, and the world to function more efficiently.}
Increasing the transparency of government spending specifically is in high demand from the public. For instance, in the survey publication~\cite{localtransparency}, \enquote{Public access to records is crucial to the functioning government} was rated with a mean of 4.14 (1 = disagree completely, 5 =  agree completely).
Open spending data can reduce corruption by increasing accountability and strengthening democracy because voters can make better informed decisions.
Furthermore, an informed and trusting public also strengthens the government itself because it is more likely to commit to large projects (see~\cite{fiscaltransparency} for details).

%Both effects are proven in a study~\cite{fiscaltransparency}.
Several States and Unions are bound to financial transparency by law, such as the European Union\footnote{\enquote{2. The Commission shall make available, in an appropriate
and timely manner, information on recipients, as well as the nature and purpose of the measure financed from the budget[...]}~\cite{euregulations}} with its \emph{Financial Transparency System (FTS)}\footnote{\url{http://ec.europa.eu/budget/fts}}~\cite{martin-fts}.
Public spending services satisfy basic information needs, but in their current form they do not allow queries which go further than simple keyword search or which cannot be answered with data from one system alone.
Linked Data solves those problems by providing a unified format, a powerful query language and the possibility of integration with linked data sets from other services.

Our contribution is an RDF transformation of the OpenSpending\footnote{\url{http://openspending.org}} project which provides government spending financial transactions from all over the world and is thus suitable as a core knowledge base that can be enriched and integrated with other, more focused data sets.
Transforming OpenSpending to Linked Data and publishing it adds to and profits from the Semantic Web which offers benefits including a standardized interface, easier data integration and complex queries over multiple knowledge bases.

%Some advantages of an RDF version over a relational database are (1) querying over multiple data sets at once using interlinks and (2) an additional custom vocabulary for each of the datasets, which are very heterogeneous (see \autoref{tab:sparqlqueries} for SPARQL queries for common use cases).

 %linkage provides benefits blablabla ...
%\todo{more motivation why RDF is good users are able to get this and that, what is linked, e.g. financial transparency system, cordis -> big benefit at least it is possible to link to this and that list other data sets, such as ... e.g. the paper about greece}

%We present LinkedSpending, which provides approximately two million financial transactions from all over the world extracted from OpenSpending as \emph{Linked Open Data}.
%information needs such as browsing of data sets, visualization and simple queries but they suffer from

The structure of the paper is as follows.
\Cref{sec:motivation} motivates the work and presents use cases.
\Cref{sec:openspending} describes OpenSpending, which is the source of the data, and its statistical data model.
\Cref{sec:conversion} explains the target RDF Data Cube vocabulary and the transformation process to it.
\Cref{sec:publishing} describes, how and where the data set is published and in which way users can access the data.
%\Cref{sec:overview} gives an overview over the transformed linked dataset, describes the its schema and shows typical SPARQL queries.
\Cref{sec:overview} gives an overall view of the data sets, gives details about the licence used and describes the data sets it is interlinked to.
%This is followed by a description of usage scenarios in \Cref{sec:usagescenarios}.
\Cref{sec:relatedwork} presents related spending data sets as Linked Open Data (LOD).
%The data quality is analyzed in \cref{sec:dataquality} and
The last section discusses known shortcomings of the data sets and future work. The prefixes used throughout this publication are defined in \Cref{tab:namespaces}.
In order to save space, prefixes are used even when technically incorrect, such as in \resource{ls:berlin\_de/model}.

\begin{table}[hbtp]
\caption{Namespaces and prefixes used in the paper}
\label{tab:namespaces}
%\tiny
\scriptsize
%\footnotesize
\renewcommand{\tabcolsep}{5pt}
\begin{tabularx}{\columnwidth}{
   >{\hsize=.25\hsize}L 
   >{\hsize=1.75\hsize}L 
}
\toprule
\textbf{prefix}	&\textbf{URL}\\
\midrule
os			&\os\\
owl			&\url{http://www.w3.org/2002/07/owl\#}\\
ls			&\ls\\
lso			&\lso\\
qb			&\url{http://purl.org/linked-data/cube\#}\\
sdmxd			&\url{http://purl.org/linked-data/sdmx/2009/dimension\#}\\
dbpedia			&\url{http://dbpedia.org/resource/}\\
dbp			&\url{http://dbpedia.org/property/}\\
\bottomrule
\end{tabularx}
\vspace{-20pt}
\end{table}

%while \cref{sec:transformation} explains the Data Cube model and the transformation.
%-The purpose of LinkedSpending is explained in \cref{sec:purposeoflinkedspending}.

%While the OpenSpending API is enough to satisfy basic information needs such as browsing of data sets, visualization and simple queries, Linked Data provides many more benefits such as complex queries and the possible integration of other datasets such as the FTS and
%OpenSpending.org is a an open platform that provides public finance data from governments around the world.
%At the time of this writing,

%We converted this data to RDF Data Cubes using a Java program which can be called again should data be added or existing data updated.

\section{Motivation}\label{sec:motivation}
%\section{Usage Scenarios}\label{sec:usagescenarios}
In a time of globalization,
%in the finance sector
financial data becomes an international network.
RDF data with its linked nature supports a representation that takes this network nature into account.
As a machine interpretable format, it lowers the access barrier for application developers.
For instance, generic Linked Data tools such as OntoWiki, CubeViz and Facete provide end users with the means to explore the data and discover new insights.

%\textbf{Community Empowerment:}
%network effect 
%reduce corruptions
% corruption by increasing accountability and strengthening democracy because voters can make better informed decisions.
%Furthermore, an informed and trusting public also strengthens the government itself because it is more likely to commit to large projects.
%hub

%domain specific tool on dataset

%generic tool

%\paragraph{Internationalization Support}

%Apart from the information needs that can already be satisfied using the source data, there are usage scenarios that are easier or only possible with LinkedSpending.

%\subsection{Economic Analysis}
\paragraph{Economic Analysis}
LinkedSpending is represented in Linked Open Data, which facilitates data integration.
Currencies from DBpedia and countries from LinkedGeoData are already integrated.
Financial data offers further integration candidates, such as political or other statistical, policy-influencing data such as health care.
This allows queries such as query 7 in \autoref{tab:sparqlqueries}, which asks for data sets with currencies whose inflation rates are greater than \SI{10}{\percent}.
%\todo{Ev. noch ein weiteres Beispiel um das noch zu untermauern und etwas mehr Substanz in die Subsection zu bringen - auch im Vergleich zu den folgenden beiden Subsections.}.

LinkedSpending can also be used to compute economic indicators across several data sets.
A possible indicator is a country's spending on education per person where the population size can be taken from the LinkedGeoData countries linked from one or more budget data sets.
One such data set is \dataset{ugandabudget}, which contains the Uganda Budget and Aid to Uganda, 2003--2006.
LinkedSpending serves as a hub for the integration of those data sets and their provenance information.
More data sets can be integrated with similarity-based interlinking tools such as LIMES~\cite{NGON11}.
%\footnotetext{and its web interface SAIM\cite{Lyko2013}}%, available at \url{http://saim.aksw.org}}

%\subsection{Putting Spending Amounts into Perspective}
%\subsection{Finding and Comparing Relevant Datasets}
\paragraph{Finding and Comparing Relevant Data Sets}

Government spending amounts are often much higher than the sums ordinary people are used to dealing with but even for policy makers it is hard to understand whether a certain amount of money spent is too high or normal.
Comparing data sets and finding those which are similar to another one helps separating common values from outliers which should be further investigated.
For example, if another country has a similar budget structure but spends way less on health care with a similar health level,
it should be investigated whether that discrepancy is caused by inherent differences such as different minimum wages or a different climate or if it is due to preventable factors such as inefficiencies or corruption.
While OpenSpending provides several hundreds of data sets which can be searched and it allows browsing and visualization of any single one, it does not provide a comparison function between data sets.
Because of the mechanism to identify equivalent properties (see \autoref{sec:conversion}), SPARQL queries can compare different data sets, e.g.~between similar structures in different countries.
Query 9 in \autoref{tab:sparqlqueries} shows a simple query to detect data sets which are most similar to any particular data set.
This is done by calculating the number of common measures, attributes and dimensions.
\iffalse
\subsection{Increased E-Government Transparency}
While OpenSpending makes government spending data available, the data can be made more understandable by providing additional knowledge about the described entities and by facilitating access to the data by non-expert users.
\paragraph{Additional Knowledge}
Through integration with other knowledge bases, users can for example learn about the country and currency a data set refers to at the same time as they query the data. \todo{Scheint mir nicht ideal formuliert. Während des Queries etwas lernen, kann man höchstens wenn der Endpoint sehr langsam ist. Ev. "while they browse and explore the data", falls das gemeint ist? Es wäre auch gut, wenn 7.1 und 7.3 nicht beide identisch Integration von country und currency erwähnen, da es sonst so wirkt, als ob es nicht viele andere Möglichkeiten gibt und man versucht den gleichen Punkt mehrfach an die Reviewer zu verkaufen.}
This helps interpreting the values and constructing helpful indicators. For example, the budget countries spend on health care varies wildly but less so when it is divided by the population number.

\todo{ich habe das qa jetzt rausgenommen da mir kein besonders guter use case eingefallen ist und die aus den example sparql queries zu kompliziert wären für eine erfolgreiche beantwortung}
\fi
\iffalse
\paragraph{Question Answering}
The OpenSpending data is available through numerous interfaces such as a JSON API, a keyword search and graphical visualizations.
Linked Open Data however allows the additional usage of Question Answering applications in order to to query the data in a way which is both more intuitive for non-experts than SQL and more expressive than keyword queries, which suffer from the inability to express relations and context, or faceted search, which only allows certain types of search restrictions.
Question Answering approaches, such as TBSL~\cite{unger2012template}, can thus be adapted to the spending domain and then allow not only computer science experts, but journalists or just ordinary people to know more about it more intuitively using natural, full-text sentences.
\todo{Der Paragraph sollte noch etwas angepasst werden um zu zeigen, warum Question Answering gerade hier günstig ist, z.B. eine konkrete Abfrage oder man probiert tatsächlich TBSL mit den Daten aus. Aktuell liest sich der Paragraph eher wie eine Begründung warum Question Answering über RDF allgemein sinnvoll ist (was zweifelslos der Fall ist, aber die Relevanz für das Paper noch nicht begründet).}
\fi

%\section{Source for the Data and Purpose of the Transformation}\label{sec:sourceforthedataandtopiccoverage}
%\section{Purpose of LinkedSpending}\label{sec:purposeoflinkedspending}
% viellecht besser passende beispiele nehmen, also anstelle von Kontinenten Gruppen und Hauptgruppen nehmen
%For example the coded property \url{:continents} could have the range \url{:Afrika}, \url{:America}, \url{:Asia}, \url{:Australia} and the child property \url{:countries} ... (spezielle properties für jeden kontinent?)
%This parent-child relationship is different from rdfs:subClassOf because the values are disjunctive but instead the childrens values are parts of the parents values.

%grouped in slices data sets 
%structure defined by datastructure definitions

%At the time of this writing, OpenSpending contains 217 different data sets.
%While they share the general \emph{budget} domain, they still have slightly different contexts and are described in different vocabularies.

\iffalse
2-3 beispiele

statistik darüber bringen

manuelles proof über die qualität

am ende die datacubes nehmen

data quality dashboard vom data cube nehmen
\fi

\section{OpenSpending Source Data}\label{sec:openspending}
OpenSpending\footnote{\url{http://openspending.org/}} is a project which aims to track and analyze public spending worldwide and, at May 2014, contains more than 25 million financial entries in \numberofopenspendingdatasets{} data sets\footnote{As some of them contain errors, the number of LinkedSpending data sets is slightly smaller.}.
Data Sets can be submitted and modified by anyone but they have to pass a sanity check from the OpenSpending Data Team which also cleans the data before publishing.\footnote{\url{http://community.openspending.org/contribute/data/}}
OpenSpending hosts transactional as well as budgetary data with a focus on government finance.\footnote{\url{http://community.openspending.org/help/guide/en/financial-data-types/}}
It contains this data in structured form stored in database tables and provides searching and filtering as well as visualizations and a JSON REST interface.
The data sets differ in granularity and type of accompanying information, but they share the same meta model.
%, e.g. from the FTS or CORDIS\footnotemark.
\subsection{The Data Cube Model}%\label{sec:rdfdatacubemodel}
The domain model of OpenSpending is a \emph{data cube} (also \emph{OLAP cube}, \emph{hypercube}), which represents multi-dimensional statistical observations.
Each cell corresponds to an observation (an instance of spending or revenue) that contains measurements (e.g. the amount of money spent or received).
The context of the measurement is provided by the \emph{dimensions} like the purpose, department and time of a spending item and optionally by \emph{attributes}, which further describe the measured value, e.g., the unit of the measurement.
%Apart from the fixed data cube meta model, the structure of each dataset is completely up to the creator.
%The budget data sets for example have the dimensions \emph{time} and \emph{department}, the measure \emph{amount spent} and the attribute \emph{currency} with the value \emph{euro}.
\begin{figure}[h]
\footnotesize
\begin{lstlisting}[language=json]
"sub-programme": {
 "label": "Sub-programme",
 "type": "compound",
},
"amount": {
    "datatype": "float",
    "label": "Total",
    "type": "measure",
}
\end{lstlisting}
\caption{simplified excerpt of an OpenSpending \emph{model}} 
\label{fig:jsonmodel}
\end{figure}

\begin{figure}[h]
{\footnotesize
\begin{lstlisting}[language=json]
"sub-programme": {
 "label":"Security and safeguarding liberties",
 "html_url":"http://openspending.org/eu-budget/sub-programme/security-and-safeguarding-liberties",
 "name":"security-and-safeguarding-liberties"
},
"html_url": "http://openspending.org/eu-budget/entries/017dfcb58d05671ef9eb5a9f77fef39c8b14150c",
"amount": 41.2
\end{lstlisting}
}
\iffalse
"main-programme": {

    "description": null,
    "label": "Main programme",
    "facet": true,
    "key": true,
    "attributes": {
        "name": {
            "column": "StringCAT1",
            "datatype": "id",
            "default_value": ""
        },
        "label": {
            "column": "StringCAT1",
            "datatype": "string",
            "default_value": ""
        }
    },
    "type": "compound",
    "dimension": "main-programme"
},
"main-programme": {
 "label":"Citizenship, Freedom ,Security and Justice",
 "html_url":"http://openspending.org/eu-budget/main-programme/citizenship-freedom-security-and-justice",
 "name":"citizenship-freedom-security-and-justice"
},

\fi
\caption{simplified excerpt from an OpenSpending \emph{entry}}
\label{fig:jsonentry}
\end{figure}

\Cref{fig:jsonmodel} shows an excerpt from the model of the OpenSpending data set \emph{eu-budget} with the dimension \emph{sub-programme} and the measure \emph{amount}.
\Cref{fig:jsonentry} shows the an entry that contains the actual values for the dimension and the measure of the observation.

\subsection{Problems}
While the data is well-structured and thus suitable for conversion without data cleaning or extensive preprocessing, it still poses problems that need to be taken into account:
\begin{inparaenum}
\item New data sets are frequently added (approximately 50 per month) and, less often, existing data sets are modified.
\item Some data sets do not specify a value for all properties in all observations.
\item There are properties with the same name in different data sets where it is unknown if they specify the same property.
\item Data Cube is a meta model. The deep structure of the data sets is heterogeneous and
 described only shallowly.
\item The language of literals is varying between and even within data sets but the language used is not specified.
\end{inparaenum}
Points 1 to 3 are addressed in the next section while points 4 and 5 are discussed in \cref{sec:conclusions}.

\section{Conversion of OpenSpending to RDF}\label{sec:conversion}

%The conversion from OpenSpending to Linked Data consists of the extraction of the source data and its transformation to RDF.
%Because the source data adheres to the data cube model, a conversion of the data to RDF needs an appropriate RDF vocabulary.
%The RDF DataCube vocabulary~\cite{rdfdatacube} is an ideal fit for the transformed data.
%Because the source data uses the data cube model, the RDF DataCube vocabulary~\cite{rdfdatacube} is an ideal fit for the transformed data.

\begin{table*}[hbtp]
\caption{Conversion of OpenSpending to LinkedSpending classes and instances}
\label{tab:transformation}
\scriptsize
\renewcommand{\tabcolsep}{5pt}
\begin{tabularx}{\textwidth}{
% sum of hsizes needs to be 2
   >{\hsize=.1\hsize}L 
   >{\hsize=.4\hsize}L 
   >{\hsize=.8\hsize}L 
   >{\hsize=.6\hsize}L 
   >{\hsize=.3\hsize}L          
}
\toprule % a bit unelegant: with the replacement of \url by \urlX to circumvent hyperref the line break inside urlX does not work so there are 2 spaces
&\textbf{Source URL}  		&\textbf{JSON Path}				&\textbf{LinkedSpending class}		&\textbf{LS instance scheme}\\ 
\midrule
\rom{1}  		&\osdatasetjson 		&						&\class{qb:DataSet}			&\lsdataset\\
\rom{2}			&\osmodel 			&						&\class{qb:DataStructureDefinition}	&\dsd\\
\midrule
\multirow{3}*{\rom{3}} 	&\osmodel			&\texttt{\$.mapping.*}				&\class{os:\{Country,Time\}Component} \class{Specification}
													or \class{qb:ComponentSpecification}	&\urlX{lso:}\emph{propertyname} \urlX{-spec}\\
\midrule
\rom{4}			&\multirow{4}*{\osmodel}	&\texttt{\$.mapping.*[?(@.type="compound")]}	&\class{qb:DimensionProperty}		&\multirow{4}*{\urlX{lso:}\emph{propertyname}}\\
\rom{5}			&				&\texttt{\$.mapping.*[?(@.type="date")]}		&\class{qb:DimensionProperty}	&\\
\rom{6}			&				&\texttt{\$.mapping.*[?(@.type="measure")]}	&\class{qb:MeasureProperty}		&\\
\rom{7}			&				&\texttt{\$.mapping.*[?(@.type="attribute")]}	&\class{qb:AttributeProperty}		&\\
\midrule
\rom{8} 		&\osdataset\urlX{/entries.json} 		&\texttt{\$.results[*].dataset}	&\class{qb:Observation}				&\urlX{ls:}\urlX{observation-} \emph{dataset name}-\emph{hashvalue}\\
\bottomrule
\end{tabularx}
\end{table*}
%\iffalse
%\begin{sidewaystable}
%{\setlength{\tabcolsep}{0.5em}
%\scriptsize
%\begin{tabulary}{0.84\textwidth}{lllLp{3cm}p{3cm}}
%\toprule
%			&\textbf{Source URL}  		&\textbf{JSON Path}				&\textbf{LinkedSpending class}		&\textbf{OpenSpending URL}				&\textbf{LinkedSpending URI}\\ 
%\midrule
%\rom{1}  		&\osdatasetjson 		&						&\class{qb:DataSet}			&\osdataset						&\lsdataset\\
%\rom{2}			&\osmodel 			&						&\class{qb:DataStructureDefinition}	&\osmodel						&\dsd\\
%\midrule
%\multirow{3}*{\rom{3}} 	&\osmodel			&\texttt{\$.mapping.*}				&\class{os:{Country,Time}ComponentSpecification}
%													or \class{qb:ComponentSpecification}	&none							&\lsdataset\url{/}\emph{propertyname} \url{/component} \url{Specification}		\\
%\midrule
%\rom{4}			&\multirow{4}*{\osmodel}	&\texttt{\$.mapping.*[?(@.type="compound")]}	&\class{qb:DimensionProperty}		&\multirow{4}*{\osdataset\url{/}\emph{propertyname}}	&\multirow{4}*{\lsdataset\url{/}\emph{propertyname}}\\
%\rom{5}			&				&\texttt{\$.mapping.*[?(@.type="date")]}		&\class{qb:DimensionProperty}	&							&\\
%\rom{6}			&				&\texttt{\$.mapping.*[?(@.type="measure")]}	&\class{qb:MeasureProperty}		&							&\\
%\rom{7}			&				&\texttt{\$.mapping.*[?(@.type="attribute")]}	&\class{qb:AttributeProperty}		&							&\\
%\midrule
%\rom{8} 		&\osdataset\url{/entries.json} 		&\texttt{\$.results[*].dataset}	&\class{qb:Observation}			&\osmodel\url{/entries/} \emph{hashvalue}			&\lsdataset\url{/model/entries/}\emph{hashvalue}\\
%\bottomrule
%\end{tabulary}
%\caption{Transformation of OpenSpending entities to LinkedSpending classes}}
%\label{tab:transformation}
%\end{sidewaystable}
%\fi
%\todo{\autoref{tab:transformation} source uris take the ones where i get them from or where the source uri is that only describes it? should be both acutally}

%\paragraph{The RDF DataCube vocabulary}
\paragraph{}
The RDF DataCube vocabulary~\cite{rdfdatacube}, i.e.~an RDF variant of the previously explained data cube model, is an ideal fit for the transformed data. 
\begin{figure}[h]
\includegraphics[width=\columnwidth]{img/rdfdatacube_shrinked.pdf}
\caption{Used RDF DataCube concepts and their relationships\protect\footnotemark}
\label{fig:rdfdatacube}
\end{figure}
\footnotetext{Simplified version of the structure described in ~\cite{rdfdatacube}.}

First and foremost, this vocabulary provides the backbone structure for every LinkedSpending data set, see \cref{fig:rdfdatacube}.
Each data set is represented by an instance of \class{qb:DataSet} and an associated instance of \class{qb:DataStructureDefinition} which includes \emph{component specifications} (see \cref{fig:modellingexample} for an example).
Each component specification is associated to a \emph{component property} which can be either a \emph{dimension}, an \emph{attribute} or a \emph{measure}.
Commonly used concepts are specified in the model of the \emph{Statistical Data and Metadata eXchange (SDMX)} initiative\footnote{\url{http://sdmx.org}}.
%\Cref{fig:modellingexample} shows, how
%example, a reference area can make use of \resource{sdmxa:refArea}.
The RDF Data Cube vocabulary  is supported by the LOD2 Statistical Office Workbench\footnote{\url{http://demo.lod2.eu/lod2statworkbench}} which is part of the Linked Data Stack (an advanced version of the the LOD2 Stack~\cite{Auer+ISWC-2012}).
The workbench includes a DataCube validator, a split and merge component and a CKAN Publisher.
The OntoWiki~\cite{ontowiki_www}, which manages several parts of the the Linked Data Lifecycle~\cite{Auer+ISWC-2012}, such as Storage/Querying and Search/Browsing/Exploration offers a CSV import plugin for the format as well as a faceted RDF Data Cube browser, CubeViz.
Data cubes may contain slices, which are presets for certain dimension values, effectively selecting a subset of a cube. Users may create and visualize their own slices using the OntoWiki CubeViz plugin. Furthermore, the RDF DataCube vocabulary allows the persistence of slices which is used to represent preconfigured slices from OpenSpending.

% the Linked Data Stack~\cite{Auer+ISWC-2012}
% which others? ORE - Ontology Repair and Enrichment ?

\begin{figure}
\footnotesize
\begin{lstlisting}[language=ttl]
ls:berlin_de
 rdf:type	qb:DataSet;
 rdfs:label	"Berlin Budget";
 dc:source	os:berlin_de;
 qb:structure	ls:berlin_de/model;
 qb:slice	ls:berlin_de/views/nach-einzelplan.

ls:berlin_de/model
 rdf:type qb:DataStructureDefinition;
 qb:component
   lso:CountryComponentSpecification,
   lso:DateComponentSpecification,
   lso:Einzelplan-spec,
   lso:amount-spec.

lso:CountryComponentSpecification 
 rdf:type qb:ComponentSpecification;
 rdfs:label "country";
 qb:attribute sdmxa:refArea; 
 qb:componentRequired "false"^^xsd:boolean;
 qb:componentAttachment
   qb:DataSet,qb:Observation.
\end{lstlisting}
\caption{RDF DataCube vocabulary modelling excerpt of data set \resource{berlin\_de} (some properties and values omitted).}
\label{fig:modellingexample}
\end{figure}

\paragraph{Transformation}
All of the OpenSpending data sets describe observations referring to a specific point or period in time and thus undergo only minor changes.
New data sets however, are frequently added. Because of this, the huge number of data sets and their size, an automatic, repeatable transformation is required.
This is realized by a program\footnote{written in Java, available as open source at \url{https://github.com/AKSW/openspending2rdf}} which fetches a list of data sets on execution and only transforms the ones who are not transformed yet. Each data set is transformed separately.
\autoref{tab:transformation} shows for each class used by LinkedSpending, at which URL (abbreviated using the prefixes from \cref{tab:namespaces}) the information used to create the instances of those classes is found. % cref somehow doesnt work with this table
In case there are multiple instances described at one URL, a \emph{JSON path}\footnotemark{} expression is given, that locates the corresponding subnodes.
\footnotetext{\emph{JSON path} (\url{http://code.google.com/p/json-path/}) is a a query language for selecting nodes from a JSON documents, similar to XPath for XML}
Finally, the table contains the patterns that describe resulting LinkedSpending URLs.
For example, the OpenSpending URL \resource{os:berlin\_de/model} contains the node \texttt{\$.mapping.amount} which has a type value of \enquote{attribute} and is, thus, transformed to the OpenSpending instance \resource{lso:amount} of the class \resource{qb:AttributeProperty}.

Equivalent component properties (dimensions, attributes and measures) are identified as follows:
A configuration file optionally specifies the mapping of data set and property name to an entity in the LinkedSpending ontology.
By default, the property URI is derived from the property name.
Properties with the same name in different data sets not having a mapping entry that states otherwise are assumed to represent the same concept and thus given the same URL.\footnotemark
\footnotetext{Although that has the possibility of mismatches, such a mismatch has not been spotted yet. Still, evaluating and, if necessary, improving the automatic matching is part of future work.}

%\subsection{Use of Established Vocabularies}
\paragraph{Use of Established Vocabularies}
%\textbf{Use of Established Vocabularies:} 
In addition to the standard vocabularies, RDF, RDFS, OWL and XSD, the DCMI vocabulary is used for source and generation time metadata.
The data sets are modelled, first and foremost, according to the RDF Data Cube vocabulary, % (see \cref{sec:openspending}),
which specifies the structure of a data cube.
LinkedSpending follows the RDF Data Cube recommendation to make heavy use of the SDMX model for measures, attributes and dimensions.%, among others, sponsored by the United Nations Statistics Division, the International Monetary Fund and the World Bank.
The data sets are very heterogeneous but there are some properties which are commonly specified and thus modelled with established vocabularies.
The year and date, a data set and an observation refers to, respectively, is expressed by \resource{sdmx-dimension:refPeriod} and XSD.
%other than the usage of time and dates, which are expressed  currencies and countries.

Currencies are taken from DBpedia~\cite{dbpedia-swj} and countries are represented using the vocabulary of LinkedGeoData~\cite{SLHA11}, a hub for spatial linked data.
Some amount of data is imported from LinkedGeoData countries and DBpedia currencies. Because of the limited number of countries and currencies, and properties values imported per country and currency, the amount of data is too small to consider federated querying.
As most countries and currencies are stable in the medium term, this data needs to be updated only infrequently.
\paragraph{Interlinking}
%\textbf{Interlinking:}
There are two possibilities to align entities to another vocabulary: 1) to use the entities directly and 2) to create an own RDF resource with interlinks, like \resource{owl:sameAs}, to that vocabulary.
We generally preferred the first approach because a higher amount of reuse provides easier integration, better understandability and tool support.
%We generally preferred to use entities of a vocabularies directly, such as the currencies and countries, because this is a more elegant and less convoluted approach as long as the used vocabularies are stable and exactly correspond to the required meaning.
%, as long as the entities are always in the object, or property position.
% wir brauchen nicht extra begründen warum man subjects verlinken muss um dann zu sagen, dass es sowieso keine observations gibt, die man verknüpfen kann (da fragt sich der leser warum wir das schreiben)
%Using an entity of another vocabulary in the subject position is problematic, however, because, amongst other things, those additional triples are not reachable by standard Linked Data URI dereferencing.
%Because of this, the interlinks between observations of the data sets to observations in other data cubes need to be aligned by interlinking and not by replacement.
% Observations are only equal to others however, when they describe exactly the same transaction and overlapping data sets from other sources have not been found.
% While there are usually no interlinking targets for \emph{sameAs} links to other observations, there are many possibilities for interlinks between data sets or dimension values (which are always URIs, mostly with labels) and concepts they refer to.
While we did not find \emph{sameAs} link targets on observation level, i.e.~exactly the same statistical observations described in other data sets, there are many possibilities for interlinks between data sets or dimension values and concepts they refer to.
Using the labels of those data sets and dimension values, it is possible, for example, to link values of the dimension \enquote{region} of a federal budget, and thus indirectly also the observations which use those values, to the cities in DBpedia or LinkedGeoData whose labels are contained in the label of the region value URI.

%\todo{Wenn zwei Datensätze sich z.B. auf Städte beziehen, könnte man vermutlich schon mehr Sachen verknüpfen. Das Problem bei dem Absatz ist, dass Du hier sagst, dass es quasi keine Linkmöglichkeiten gibt, aber unter 7.1 betonst, dass man mit mehr Links noch mehr Analysen durchführen kann.}

%, such as \ls\url{/berlin\_de}, 
%, can browse single URLs as HTML through the Pubby\footnote{\url{http://wifo5-03.informatik.uni-mannheim.de/pubby/}} interface by dereferencing the URIs, such as \ls\url{/berlin\_de}, in a browser.

%The user can browse single URLs as HTML through the Pubby\footnote{\url{http://wifo5-03.informatik.uni-mannheim.de/pubby/}} interface by dereferencing the URIs, such as \ls\url{/berlin\_de}, in a browser.

%Because mass processing such as interlinking requires computationally expensive queries against the SPARQL endpoint, an RDF dump\footnote{\ntriples} is also available. \todo{Satz würde ich eher weglassen bzw. verschieben. Es ist recht klar, dass es einen Dump gibt und es passt nicht in die Section.}
%\Cref{tab:technicaldetails}

\paragraph{Error Handling}
The OpenSpending API lists \numberofopenspendingdatasets{} data sets with \numberofdatasets{} of them having a LinkedSpending equivalent.
The discrepancy is caused by loss in several stages.
To prevent timeouts and to reduce the impact of disrupted connections, the source data set is downloaded in several parts with a maximum number of entries.
These parts are then merged so that each file corresponds to exactly one data set.
Data sets without observations are removed and the remaining data sets are transformed, noting the missing values for all component properties.
If the first 1000 values are all missing, the transformation is aborted, otherwise a \resource{lso:completeness} value $c = \frac{|\textnormal{existing values}|}{|\textnormal{observations}||\textnormal{component properties}|}$ is attached to the data set.
Besides empty or non-existing data sets, there were no other types of error observed.
%here are however several cases of component properties with the same name which raises the problem of determining equivalent component properties. The chosen approach is to regard as equal all properties with exactly the same name.
%A%ll data sets, with the exception of ones where the first 1000 expected values are all missing, are now fully converted and each dataset now has a property “” which has an assigned float value measuring 1-%(missing number of values)/(expected number of values).

>>>>

% add. options: [seceqn,secthm,crcready,onecolumn]
\documentclass[sw]{iosart2x}

%\usepackage{dcolumn}
%\usepackage{endnotes}
\usepackage{booktabs}
\usepackage{csquotes}
%\usepackage{graphicx}
\usepackage{todonotes}
\setlength{\marginparwidth}{2.1cm}% enough space for todonotes
%\usepackage{listings}
%\lstset{language=SPARQL,breaklines=true}
\usepackage{cleveref}

@@

% Two or more authors:
\author[A]{\inits {K.}\fnms{Konrad} \snm{Höffner}\ead[label=e1]{konrad.hoeffner@imise.uni-leipzig.de}%
\thanks{Corresponding author. \printead{e1}.}},
\author[A]{\fnms{Franziska} \snm{Jahn}\ead[label=e2]{franziska.jahn@imise.uni-leipzig.de}}
\author[A]{\fnms{Birgit} \snm{Schneider}\ead[label=e3]{birgit.schneider@imise.uni-leipzig.de}}
\author[A]{\fnms{Anna} \snm{Lörke}\ead[label=e4]{anna.loerke@imise.uni-leipzig.de}}
\author[A]{\fnms{Thomas} \snm{Pause}\ead[label=e5]{thomas.pause@imise.uni-leipzig.de}}
\author[A]{\fnms{Alfred} \snm{Winter}\ead[label=e6]{alfred.winter@imise.uni-leipzig.de}}
\runningauthor{}
\address[A]{Institute for Medical Informatics, Statistics and Epidemiology (IMISE),
\orgname{University of Leipzig}, \cny{Germany}\printead[presep={\\}]{e1,e2,e3,e4,e5,e6}}
%Medical Informatics, Management of Health Information Systemsi
%Härtelstraße 16--18, D-04107 Leipzig

\begin{abstract}
Abstract text.
\end{abstract}

\begin{keyword}
\kwd{information management, information systems, hospital information management}
\end{keyword}

\end{frontmatter}

\begin{table}
\caption{}
\label{tab:namespaces}
\begin{tabular}{ll}
\toprule
\textbf{x}	&\textbf{y}\\
\midrule
\bottomrule
\end{tabular}
\end{table}

\section{Background}
\todo{not Konrad: write background of meta model}

\section{Architecture}
The \textbf{Semantic Network of Information Management in Hospitals} (\enquote{Krankenhaus} in German) is an OWL 2 DL ontology that consists of a central meta model and modular subontologies based on the meta model.

\section{Data Set Description}

\section{Application}

\section{Acknowledgments}
We thank the publishers ... and ... for granting us permission to publish the subontologies based on the textbooks [][][] under an open license.
The SNIK project is supported by the DFG (German Research Foundation) under the grant numbers 1605/7-1 and 1387/8-1.
\nocite{*} 
\bibliographystyle{ios1}
\bibliography{paper}
